SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-11-19 abs.tex
Mostly editing for English, redundancy, some technical changes, and redundancy.

\begin{abstract}
Hardware prefetching has been introduced in modern
processors as a way to hide cache latencies.  An efficient
prefetcher should be able to identify complex memory access
patterns during a program execution.  This enables it to fetch a
block ahead of its demand access, potentially saving a cache miss.

In this paper, we introduce perceptron learning to help make
this prefetching decision efficiently. The perceptron layer acts 
as a throttler to filter out the unnecessary prefetches recommended 
by the underlying prefetcher.  We have also explored a range 
of features that can be used to train the perceptron layer.  
Our results show that NATCH
improves performance on the memory intensive subset of the SPEC
2017 Benchmark suite by 6.84\% on single-core and by 11.9\% on
multi-core traces, as compared to state-of-the art prefetcher.  
We also demonstrate that the performance gained from using 
our efficient throttler scales better with increasing number 
of cores. 
\end{abstract}

>>>>

\begin{abstract}
Hardware prefetching has been introduced in modern processors as a way to hide
cache latencies.  An efficient prefetcher should be able to identify complex
memory access patterns during program execution. This ability enables the
prefetcher to read a block ahead of its demand access, potentially saving a
cache miss. Accurately identifying the right blocks to prefetch is essential
to achieving high performance from the prefetcher.

In this paper, we introduce Perceptron-based Prefetch Filtering to help make
this prefetching decision accurately. The perceptron layer acts as a throttler
to filter out the unnecessary prefetches recommended by the underlying
prefetcher.  We have also explored a range of features that can be used to
train the perceptron layer.  Our results show that perceptron-based filtering
improves performance on the memory intensive subset of the SPEC 2017 benchmark
suite by 6.84\% on single-core and by 11.9\% on multi-core traces, as compared
to a state-of-the art prefetcher.  We also demonstrate that the performance
gained from using our efficient throttler scales better with increasing numbers
of cores.

\end{abstract}

### 2018-11-19 back.tex
Mostly editing for English, redundancy, some technical changes, and redundancy.

In this section, we compare the existing work done in the domain of
prefetching.  We also discuss about some of the interesting memory
access patterns seen in the workloads in SPEC 2017 benchmark.  The
second half of this section deals with a background of Signature Path
Prefetcher architecture.

@@

With time newer and more intricate prefetchers were
introduced~\cite{Wenisch_Temporal_Streaming,Stealth,Feedback_Directed,Coordinated,Bandwidth_Efficient,Pacman,TLB,Linearizing,Sandbox,VLDP,DoL,Domino}.
A whole class of streaming prefetchers developed on the lines of
Spatial / Temporal Memory
Streaming~\cite{Spatial_Pattern,SMS,Temporal_Instruction_Fetch,Off_Chip,STMS,SMS_JILP}.
There have been ideas for control flow speculation directed hardware prefetching~\cite{BFetch,MTBFetch}.

Most modern prefetchers aim at identifying complex memory access
patterns in a given application.  This allows them to capture some of
the irregular patterns seen in pointer chasing data-structures.
Offset based prefetchers are a generalization of the next-line
prefetcher.  Some of the prominent prefetchers exploiting this idea
are Sandbox Prefetcher~\cite{Sandbox} and the Best Offset Prefetcher~\cite{BOP}. 
Another class of prefetchers is lookahead Prefetchers.
Such prefetchers try to speculate deep into the application's memory
access trail.  These include Path Confidence based Lookahead
Prefetching~\cite{SPP} and Kill the Program Counter~\cite{KPC}.

A related line of work also explores cache replacement - insertion
policies and dead block prediction, with an eye towards
prefetching~\cite{DB_Pred,Cache_Burst,KPC,Harmony}.  Nori
\textit{et.al.} introduced prefetching with respect to runtime
criticality~\cite{CATCH}.

\textit{Machine Learning and Prefetching:} Peled \textit{et.al.}
introduce interesting ideas for on-line Reinforcement Learning and
dynamically scaling the magnitude of feedback given to the baseline
prefetcher~\cite{Semantics}. The biggest challenge here is that it
relies on compiler support for getting features to build the context.

The paper by Liao \textit{et. al.} focuses on prefetching for data
center applications~\cite{Datacenter}. They use offline Machine
Learning Algorithms like SVMs and Logistic Regressions to do
parametric search for an optimal prefetcher configuration.

%% NOTE: The below paper is literally the same idea as ours. But poor implementation and no good results. How much to talk about this paper?
Finally, the paper on Data Cache Prefetching with Perceptron
Learning~\cite{BadPerc} talks about the idea of two step prefetching.
The first step is an existing baseline prefetcher while the second
step is a perceptron based throttler.  The paper has lots of
limitations in terms of design and implementation.  That reflects in
the fact that it did not lead to any significant performance gain over
even the basic prefetchers like the Stride Prefetcher~\cite{Stride} and
the Markov Prefetcher~\cite{Markov}.

\subsection{Perceptron Learning in Architecture}
\label{Background-Perceptron}
Perceptron learning for computer architecture design has been around
for a while.  It was first popularized by Jimenez \textit{et. al.} for
doing branch prediction~\cite{Perc_Branch}. The biggest advantage of 
the perceptron model is that it can adjust its weight dynamically 
to increase / decrease the importance given to each feature - branch 
hostories in this case.
The structure that we use is derived from the model introduced in
Piecewise Linear Branch Prediction~\cite{Piece_Linear}.  Here the
feature itself is used to index into a hash of perceptron weights.
This form of indexing makes sure that the hyper-plane learned by the
perceptron weights is able to differentiate between linearly
inseparable outcomes.  \textit{<Is it true? Not sure>}

Perceptron Learning for Reuse Prediction~\cite{Perc_Reuse} extended the
hash perceptron architecture to perform prediction in context of cache
replacement.  Perceptron layer can learn to correlate the cache
replacement behaviour with a variety of features. The features
introduced in that paper were derived from Program Counters of the
current and the last few instructions and from the current memory 
access address.  This led to a
prediction mechanism that is highly accurate, has quick learning rate
and is adaptable to changing program behaviour.

The work was further extended by Multiperspective Reuse
Prediction~\cite{Multiperspective}, which introduced a varied set of
parametric features.  Although design space exploration for feature
and parameter selection was a non-trivial task, once fixed, it could
give a cache replacement predictor highly suitable for the given set
of applications.

\textit{Deviations From Actual Perceptrons:} Traditionally, a
perceptron prediction involves multiplying the vector of input
features: F\textsubscript{1xN}, with the corresponding weight vector:
W\textsubscript{Nx1} in a dot product fashion to obtain the sum
y\textsubscript{out}.  Here what we use is a perceptron-like
structure.  The feature is used to hash into the weight of perceptrons
and the retrieved weights are added straight away.  This can be seen a
dot product of $\vec{1}$\textsubscript{1xN} with W\textsubscript{Nx1}.
This way, the perceptron algorithm doesn't introduce any
multiplication operations in the inference or training process.  Hence
what we adapt in this work, is a perceptron-like learning algorithm as
it involves the same principle involved in perceptron inferencing and
training.  For simplicity, we will call our implementation as
perceptron implementation.

\textit{Perceptron Update Rule:} In all the above implementations, a
uniform perceptron update principle is being followed.  The weights
need to be updated if the prediction was wrong or the predicted sum
does not cross a certain threshold.  After a certain training period,
these weights are proportional to the probability of the outcome being
true.  This training threshold makes sure that the perceptrons are
trained till a certain level of confidence and yet they are not
over-trained to the given set.  Care is taken that the weights
saturate at certain positive and negative values so that they remain
confined to the bit-width.  The same rules will be applied in the
discussion done in Section \ref{Design-Perceptron} on NATCH learning algorithm.

\subsection{Prefetching with SPEC 2017}
\label{Background-SPEC2017}
\textit{<Discuss SPEC 2017 traces here>
<What are the possible graphs / analysis tools we can use for SPEC traces?>}

@@

@@

@@

\textbf{Prefetch Filter:} The PF is a 1024 entry 1-way associative table 
which keeps a record of
last few entries prefetched.  This proves useful for updating the prefetcher 
states if a tracked prefetch leads to a demand hit or a cache eviction.

@@

`D' is the lookahead depth.  For d = 1, when SPP is in
non-speculative mode, P\textsubscript{0} can be thought of as 1. 
The final P\textsubscript{d} is thresholded against prefetching-threshold
(T\textsubscript{p}) to decide whether to prefetch or not.  If yes,
then P\textsubscript{d} is thresholded against a numerically bigger
fill-threshold (T\textsubscript{f}) to decide whether the prefetch
should be sent to L2 Cache (high confidence prefetch) or Last Level
Cache (low confidence prefetch)

@@

\subsection{Case for an On-line Throttler}
\label{Background-Case}
As compared to some of the other state of the art prefetchers [BOP],
SPP is a less aggressive prefetcher.  In the single core
environment, this gives BOP an edge as there is no resource contention
among the different cores.  Hence more aggressive prefetching is bound
to prove beneficial.  As we increase the core count, we observe that
SPP starts outperforming rest of the prefetchers.  This can be
attributed to the fact that each prefetch suggested by SPP is a
carefully calculated one and that prevents cache pollution.
\textit{<Figure showing the variation of SPP vs BOP wrt core count>}
For 4-core applications, SPP suggests XX\% fewer prefetches than BOP
and yet leads to higher IPC.

The above analysis shows that with a more careful throttling
mechanism, any prefetcher like SPP can be tuned to become much more aggressive, leading to
increased coverage.  The onus of maintaining the accuracy now falls on
the independent throttler.  To test the hypothesis, we tuned down SPP
to the minimum possible threshold.  Only by doing this we managed to
increase the prefetch suggestions made by SPP, by XX\%.
\textit{<Comparison of SPP-Unleased with SPP / BOP>} Obviously this
came at a cost of increased DRAM traffic and cache pollution.

Moreover, the on-line confidence mechanism introduced in SPP was very
rudimentary.  It was based on taking a ratio C\textsubscript{d} =
C\textsubscript{delta} / C\textsubscript{sig} as explained previously. This 
confidence was used to make the decision whether to prefetch or
not to prefetch; and which level to prefetch.  While this
approximation was shown to work in the original implementation, we believe that a
better form of generalised on-line decision making was possible.  
Hence, it was
necessary to build a robust and adaptable learning mechanism to accept
/ reject the prefetch suggestions; and to decide the fill level (L2
Cache vs Last Level Cache).

>>>>

In this section, we compare the existing work done in the domain of
prefetching.  We also discuss interesting memory access patterns seen in the
workloads in SPEC 2017 benchmark.  The second half of this section gives
background for the Signature Path Prefetcher (SPP) architecture.

@@

With time, newer and more intricate prefetchers were
introduced~\cite{Wenisch_Temporal_Streaming,Stealth,Feedback_Directed,Coordinated,Bandwidth_Efficient,Pacman,TLB,Linearizing,Sandbox,VLDP,DoL,Domino}.
A whole class of streaming prefetchers developed on the lines of Spatial /
Temporal Memory
Streaming~\cite{Spatial_Pattern,SMS,Temporal_Instruction_Fetch,Off_Chip,STMS,SMS_JILP}.
Control-flow speculation directed hardware prefetching has also been
proposed~\cite{BFetch,MTBFetch}.

Most modern prefetchers aim at identifying complex memory access patterns in a
given application.  Thus, they capture some of the irregular patterns seen in
pointer chasing data-structures.  Offset based prefetchers are a
generalization of the next-line prefetcher.  Some of the prominent prefetchers
exploiting this idea are Sandbox Prefetcher~\cite{Sandbox} and the Best Offset
Prefetcher~\cite{BOP}.  Another class of prefetchers is lookahead Prefetchers.
Such prefetchers try to speculate deep into the application's memory access
path.  These include Path Confidence based Lookahead Prefetching~\cite{SPP}
and Kill the Program Counter~\cite{KPC}.

A related line of work also explores cache replacement -- insertion policies
and dead block prediction, with an eye towards
prefetching~\cite{DB_Pred,Cache_Burst,KPC,Harmony}.  Nori \textit{et.al.}
introduced prefetching with respect to runtime criticality~\cite{CATCH}.

\textit{Machine Learning and Prefetching:} Peled \textit{et.al.} introduce
interesting ideas for on-line Reinforcement Learning and dynamically scaling
the magnitude of feedback given to the baseline prefetcher~\cite{Semantics}.
The biggest challenge here is that the idea relies on compiler support for
getting features to build the context.

Liao \textit{et. al.} focus on prefetching for data center
applications~\cite{Datacenter}. They use offline machine learning algorithms
such as SVMs and logistic regression to do a parametric search for an optimal
prefetcher configuration.

%% NOTE: The below paper is literally the same idea as ours. But poor implementation and no good results. How much to talk about this paper?
% djimenez: let us not cite this paper. it is in arXiv and we are not
% obligated to cite it. as far as ISCA is concerned it does not exist. there is
% a reason it is only in arXiv and not published somewhere: it is crap.

%Finally, the paper on Data Cache Prefetching with Perceptron
%Learning~\cite{BadPerc} talks about the idea of two step prefetching.
%The first step is an existing baseline prefetcher while the second
%step is a perceptron based throttler.  The paper has lots of
%limitations in terms of design and implementation.  That reflects in
%the fact that it did not lead to any significant performance gain over
%even the basic prefetchers like the Stride Prefetcher~\cite{Stride} and
%the Markov Prefetcher~\cite{Markov}.

\subsection{Perceptron Learning in Architecture}
\label{Background-Perceptron}
Perceptron learning for computer architecture design has been around for a
while.  It was first popularized by Jim\'enez \textit{et. al.} for conditional
branch prediction~\cite{Perc_Branch}. The biggest advantage of the perceptron
model is that it can adjust its weights dynamically to increase / decrease the
importance given to each feature - branch histories in this case.  The
structure we use is derived from the model introduced for Perceptron-based
Reuse Prediction~\cite{Perc_Reuse}, refined with many additional features as
Multiperspective Reuse Prediction{Multiperspective}, and originally derived
from Hashed Perceptron branch prediction{HashedPerceptron}.  Here, a hash of
the feature itself is used to index into a table of perceptron weights.

% djimenez: this detail is unnecessary and likely to invite questions.
% reviewers might ask "what about linear separability?" but the proof is in the
% pudding.
%This form of indexing makes sure that the hyper-plane learned by the
%perceptron weights is able to differentiate between linearly
%inseparable outcomes.

Perceptron Learning for Reuse Prediction~\cite{Perc_Reuse} extended the hashed
perceptron architecture to perform prediction in context of cache replacement.
The perceptron layer learns to correlate cache replacement behaviour with a
variety of features. The features introduced in that paper were derived from
program counters of the current and the last few memory access instructions
and from the current memory access address.  This led to a prediction
mechanism that is highly accurate, has quick learning rate and is adaptable to
changing program behaviour.

The work was further extended by Multiperspective Reuse
Prediction~\cite{Multiperspective}, which introduced a varied set of
parametric features.  Although design space exploration for feature and
parameter selection was a non-trivial task, once fixed, it yields
replacement predictor highly suitable for the given set of applications.

\textit{Deviations From Actual Perceptrons:} Traditionally, a perceptron
prediction involves multiplying the vector of input features:
F\textsubscript{1xN}, with the corresponding weight vector:
W\textsubscript{Nx1} in a dot product to obtain the sum y\textsubscript{out}.
Here what we use is a perceptron-like structure.  The feature is used to hash
into the weight of perceptrons and the retrieved weights are added straight
away.  Thus, the perceptron algorithm doesn't introduce any multiplication
operations in the inference or training process.  Hence what we adapt in this
work is a perceptron-like learning algorithm as it involves the same principle
involved in perceptron inferencing and training.

\textit{Perceptron Update Rule:} In all the above implementations, a uniform
perceptron update principle is followed.  The weights need to be updated if
the prediction was wrong or the magnituded of the dot product does not exceed
a certain threshold.  After a certain training period, these weights are
proportional to the probability of the outcome being true.  This training
threshold makes sure that the perceptrons are trained until a certain level of
confidence and yet they are not over-trained to the given set.  Care is taken
that the weights saturate at the maximum positive and negative values so that
they remain confined to the bit-width.  The same rules will be applied in the
discussion done in Section \ref{Design-Perceptron} on PPF learning algorithm.

\subsection{Prefetching with SPEC 2017} \label{Background-SPEC2017}
% djimenez: we might not need to discuss the benchmarks too much as they are
% standard. we don't have a lot of room anyway.

\textit{<Discuss SPEC 2017 traces here> <What are the possible graphs /
analysis tools we can use for SPEC traces?>}

@@

% djimenez: Paul, is all this stuff accurate? SPP continues to confuse me.

@@

% djimenez: it does look like we're pushing it with all the stuff about SPP.
% Let's describe it enough to give the reader the idea that he/she has a vague
% understanding of how it works and can refer to the MICRO paper for more
% details. (I'm not saying cut anything that's already there, just don't expand.)

@@

\textbf{Prefetch Filter:} The PF is a 1024 entry 1-way associative table that
keeps a record of last few entries prefetched.  This proves useful for
updating the prefetcher states if a tracked prefetch leads to a demand hit or
a cache eviction.

@@

`D' is the lookahead depth. For $d = 1$, when SPP is in
non-speculative mode, P\textsubscript{0} can be thought of as 1. 
The final P\textsubscript{d} is thresholded against prefetching-threshold
(T\textsubscript{p}) to decide whether to prefetch or not.  If yes,
then P\textsubscript{d} is thresholded against a numerically bigger
fill-threshold (T\textsubscript{f}) to decide whether the prefetch
should be sent to L2 Cache (high confidence prefetch) or Last Level
Cache (low confidence prefetch)

@@

\subsection{Case for an On-line Throttler}
\label{Background-Case}
As compared to some of the other state of the art prefetchers [BOP], SPP is
less aggressive.  In the single-core environment, this gives BOP an edge as
there is no resource contention among the different cores.  Hence more
aggressive prefetching is bound to prove beneficial.  As we increase the core
count, we observe that SPP starts outperforming rest of the prefetchers.  This
can be attributed to the fact that each prefetch suggested by SPP is a
carefully calculated one and that prevents cache pollution.  \textit{<Figure
showing the variation of SPP vs BOP wrt core count>} For 4-core applications,
SPP suggests XX\% fewer prefetches than BOP and yet leads to higher IPC.

The above analysis shows that with a more careful throttling mechanism, any
prefetcher like SPP can be tuned to become much more aggressive, leading to
increased coverage.  The onus of maintaining the accuracy now falls on the
independent throttler.  To test the hypothesis, we tuned down SPP to the
minimum possible threshold, with the effect of increasing prefetch suggestions
made by SPP by XX\%.  \textit{<Comparison of SPP-Unleased with SPP / BOP>}
Obviously this came at a cost of increased DRAM traffic and cache pollution.

Moreover, the on-line confidence mechanism introduced in SPP was very
rudimentary.  It was based on taking a ratio C\textsubscript{d} =
C\textsubscript{delta} / C\textsubscript{sig} as explained previously. This
confidence was used to make the decision of whether to prefetch or not to
prefetch; and which level to prefetch.  While this approximation was shown to
work in the original implementation, we believe that a better form of
generalised on-line decision making was possible.  Hence, it was necessary to
build a robust and adaptable learning mechanism to accept / reject the
prefetch suggestions; and to decide the fill level (L2 Cache vs Last Level
Cache).

### 2018-11-19 conc.tex
Mostly editing for English, redundancy, some technical changes, and redundancy.

\section{CONCLUSION}
\label{Conclusion}
In this paper, we introduced neural network based learning for data
prefetching in the form of NATCH.  Perceptron acts an orthogonal
throttler to the underlying prefetch engine.  NATCH improves
performance by over XX\% over the baseline, which corresponds to YY\%
over the next best performing prefetcher.  
We believe that this is a robust and adaptable technique which can
be used to enhance any existing or upcoming prefetchers.

>>>>

\section{Conclusion}
\label{Conclusion}
In this paper, we introduced neural network based learning for data
prefetching in the form of PPF.  Perceptron acts an orthogonal
throttler to the underlying prefetch engine.  PPF improves
performance by over XX\% over the baseline, which corresponds to YY\%
over the next best performing prefetcher.  
We believe that this is a robust and adaptable technique which can
be used to enhance any existing or upcoming prefetchers.
